---
title: "Essential Statistical Data Analysis"
subtitle: "Class 1 Lab — Exploring, visualising, and constructing data"
author: "YOUR NAME GOES HERE"
date: "May 4, 2026"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    code-tools: true
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false

library(tidyverse)
library(kableExtra)

if (requireNamespace("palmerpenguins", quietly = TRUE)) {
  library(palmerpenguins)
}
```

# Lab overview

In this lab, we practise the first steps of statistical data analysis:

- inspecting a dataset;
- identifying observational units and variables;
- summarising numerical and categorical variables;
- visualising one variable;
- visualising relationships between variables;
- working with text as data;
- combining information from multiple data tables.

You can edit the code chunks directly and re-run them.

# Part 1 — Exploring a dataset

We use the `penguins` dataset from the `palmerpenguins` package.

Each row describes one penguin. The variables include species, island, bill measurements, flipper length, body mass, sex, and year.

## Lab 1 — Inspect the dataset

```{r}
penguins <- palmerpenguins::penguins

glimpse(penguins)
```

::: {.callout-tip}
### Your turn

Identify:

1. the observational unit;
2. three numerical variables;
3. three categorical variables.
:::

Write your answer here:

```{markdown}
Observational unit:

Numerical variables:

Categorical variables:
```

## Lab 2 — First rows and dimensions

```{r}
head(penguins)
```

```{r}
dim(penguins)
```

```{r}
names(penguins)
```

::: {.callout-tip}
### Your turn

How many observations are there?  
How many variables are there?
:::

Write your answer here:

```{markdown}
Number of observations:

Number of variables:
```

# Part 2 — One-variable summaries

## Lab 3 — Summarise a categorical variable

We can count how many observations belong to each category.

```{r}
penguins |>
  count(species)
```

```{r}
penguins |>
  count(island)
```

::: {.callout-tip}
### Your turn

Count the number of penguins by `sex`.
:::

```{r}
# Write your code here

```

## Lab 4 — Summarise a numerical variable

```{r}
penguins |>
  summarise(
    mean_body_mass = mean(body_mass_g, na.rm = TRUE),
    median_body_mass = median(body_mass_g, na.rm = TRUE),
    sd_body_mass = sd(body_mass_g, na.rm = TRUE)
  )
```

::: {.callout-tip}
### Your turn

Compute the mean, median, and standard deviation of `bill_length_mm`.
:::

```{r}
# Write your code here

```

## Lab 5 — Grouped summaries

We can compute summaries separately for each group.

```{r}
penguins |>
  group_by(species) |>
  summarise(
    mean_body_mass = mean(body_mass_g, na.rm = TRUE),
    median_body_mass = median(body_mass_g, na.rm = TRUE),
    sd_body_mass = sd(body_mass_g, na.rm = TRUE)
  )
```

::: {.callout-tip}
### Your turn

Compute the average `bill_length_mm` by `species`.
:::

```{r}
# Write your code here

```

# Part 3 — Visualising one variable

## Lab 6 — Histogram

A histogram is useful for visualising the distribution of a numerical variable.

```{r}
ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(
    bins = 30,
    fill = "indianred",
    color = "white",
    alpha = 0.7
  ) +
  labs(
    x = "Body mass (g)",
    y = "Frequency"
  ) +
  theme_minimal()
```

::: {.callout-tip}
### Your turn

Change the variable from `body_mass_g` to `bill_length_mm`.  
What changes in the distribution?
:::

```{r}
# Write your code here

```

Write your interpretation here:

```{markdown}
Interpretation:

```

## Lab 7 — Density plot

```{r}
ggplot(penguins, aes(x = body_mass_g)) +
  geom_density(fill = "indianred", alpha = 0.5) +
  labs(
    x = "Body mass (g)",
    y = "Density"
  ) +
  theme_minimal()
```

::: {.callout-tip}
### Your turn

Create a density plot for `flipper_length_mm`.
:::

```{r}
# Write your code here

```

## Lab 8 — Bar chart

A bar chart is useful for visualising a categorical variable.

```{r}
ggplot(penguins, aes(x = species)) +
  geom_bar(fill = "dodgerblue", alpha = 0.7) +
  labs(
    x = "Species",
    y = "Frequency"
  ) +
  theme_minimal()
```

::: {.callout-tip}
### Your turn

Create a bar chart for `island`.
:::

```{r}
# Write your code here

```

# Part 4 — Visualising relationships

## Lab 9 — Two numerical variables

A scatterplot is useful for visualising the relationship between two numerical variables.

```{r}
ggplot(
  penguins,
  aes(x = bill_length_mm, y = flipper_length_mm)
) +
  geom_point(alpha = 0.7) +
  labs(
    x = "Bill length (mm)",
    y = "Flipper length (mm)"
  ) +
  theme_minimal()
```

## Lab 10 — Adding a grouping variable

```{r}
ggplot(
  penguins,
  aes(x = bill_length_mm, y = flipper_length_mm, colour = species)
) +
  geom_point(alpha = 0.7) +
  labs(
    x = "Bill length (mm)",
    y = "Flipper length (mm)",
    colour = "Species"
  ) +
  theme_minimal()
```

::: {.callout-tip}
### Your turn

Add `facet_wrap(~ island)` to the plot.  
Does the relationship look the same on every island?
:::

```{r}
# Write your code here

```

Write your interpretation here:

```{markdown}
Interpretation:

```

## Lab 11 — One categorical and one numerical variable

Boxplots are useful for comparing a numerical variable across categories.

```{r}
penguins |>
  drop_na(species, body_mass_g) |>
  ggplot(aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot(alpha = 0.6, show.legend = FALSE) +
  labs(
    x = "Species",
    y = "Body mass (g)"
  ) +
  theme_minimal()
```

::: {.callout-tip}
### Your turn

Compare `flipper_length_mm` across `species`.
:::

```{r}
# Write your code here

```

## Lab 12 — Violin plot, boxplot, and jitter

```{r}
penguins |>
  drop_na(species, body_mass_g) |>
  ggplot(aes(x = species, y = body_mass_g, fill = species)) +
  geom_violin(alpha = 0.35, show.legend = FALSE) +
  geom_boxplot(width = 0.12, outlier.shape = NA, show.legend = FALSE) +
  geom_jitter(width = 0.08, alpha = 0.4, show.legend = FALSE) +
  labs(
    x = "Species",
    y = "Body mass (g)"
  ) +
  theme_minimal()
```

# Part 5 — Categorical association

## Lab 13 — Two categorical variables

We can count combinations of two categorical variables.

```{r}
penguins |>
  drop_na(species, island) |>
  count(species, island)
```

## Lab 14 — Row profiles

Row profiles show proportions within each row group.

```{r}
penguins |>
  drop_na(species, island) |>
  count(species, island) |>
  group_by(species) |>
  mutate(row_profile = n / sum(n))
```

::: {.callout-tip}
### Your turn

Compute row profiles for `species` and `sex`.
:::

```{r}
# Write your code here

```

## Lab 15 — Visualising categorical association

```{r}
penguins |>
  drop_na(species, island) |>
  ggplot(aes(x = species, fill = island)) +
  geom_bar(position = "fill") +
  labs(
    x = "Species",
    y = "Proportion",
    fill = "Island"
  ) +
  theme_minimal()
```

::: {.callout-tip}
### Your turn

Create a proportional bar chart for `species` and `sex`.
:::

```{r}
# Write your code here

```

# Part 6 — Text as data

Text is a useful example of non-standard data.

It does not arrive as a clean numerical matrix. We must decide what the observational units are and how to transform text into variables.

## Lab 16 — A small text dataset

```{r}
texts <- tibble::tribble(
  ~doc_id, ~field, ~text,
  1, "engineering", "The structure was tested under dynamic stress and material fatigue.",
  2, "engineering", "Mechanical systems require models of vibration, load and failure.",
  3, "engineering", "Sensors collect signals from machines under stress and vibration.",
  4, "sport", "Training load, recovery and fatigue affect athletic performance.",
  5, "sport", "Injury prevention requires monitoring intensity and recovery.",
  6, "sport", "Athletes improve performance through training, recovery and adaptation.",
  7, "economics", "Innovation and sustainability influence firms, markets and management.",
  8, "economics", "Firms compete in markets through strategy, investment and innovation.",
  9, "economics", "Management decisions affect costs, productivity and sustainability.",
  10, "humanities", "Texts, sources and contexts shape historical interpretation.",
  11, "humanities", "Archives preserve documents, memories and cultural heritage.",
  12, "humanities", "Interpretation depends on language, context and historical sources."
)

texts |>
  kbl() |>
  kable_styling(full_width = FALSE)
```

## Lab 17 — Tokenisation

Tokenisation means splitting text into smaller units, such as words.

```{r}
if (requireNamespace("tidytext", quietly = TRUE)) {
  library(tidytext)
}

tokens_raw <- texts |>
  tidytext::unnest_tokens(word, text)

tokens_raw |>
  slice_head(n = 15)
```

## Lab 18 — Remove stop words

Common words such as *the*, *and*, *of*, and *to* are often removed.

```{r}
tokens <- tokens_raw |>
  anti_join(tidytext::stop_words, by = "word")

tokens |>
  count(word, sort = TRUE)
```

## Lab 19 — Word frequencies

```{r}
tokens |>
  count(word, sort = TRUE) |>
  slice_max(n, n = 15) |>
  ggplot(aes(x = n, y = reorder(word, n))) +
  geom_col(fill = "indianred", alpha = 0.7) +
  labs(
    x = "Frequency",
    y = NULL
  ) +
  theme_minimal()
```

::: {.callout-tip}
### Your turn

Change the number of displayed words from 15 to 10.
:::

```{r}
# Write your code here

```

## Lab 20 — Word frequencies by field

```{r}
tokens |>
  count(field, word, sort = TRUE) |>
  group_by(field) |>
  slice_max(n, n = 5, with_ties = FALSE) |>
  ungroup() |>
  ggplot(aes(n, tidytext::reorder_within(word, n, field))) +
  geom_col(fill = "indianred", alpha = 0.7) +
  facet_wrap(~ field, scales = "free_y") +
  tidytext::scale_y_reordered() +
  labs(
    x = "Frequency",
    y = NULL
  ) +
  theme_minimal()
```

## Lab 21 — Document-term matrix

Many statistical methods require a rectangular data table.

A common representation is the document-term matrix:

- rows are documents;
- columns are words;
- values are word frequencies.

```{r}
document_term_matrix <- tokens |>
  count(doc_id, word) |>
  pivot_wider(
    names_from = word,
    values_from = n,
    values_fill = 0
  )

document_term_matrix |>
  select(1:min(10, ncol(document_term_matrix))) |>
  kbl() |>
  kable_styling(full_width = FALSE)
```

::: {.callout-tip}
### Your turn

How many rows does the document-term matrix have?  
What does each row represent?
:::

Write your answer here:

```{markdown}
Answer:

```

## Optional — Word cloud

```{r}
#| eval: !expr requireNamespace("wordcloud", quietly = TRUE)

library(wordcloud)

tokens |>
  count(word, sort = TRUE) |>
  with(
    wordcloud(
      words = word,
      freq = n,
      max.words = 50,
      min.freq = 1,
      random.order = FALSE,
      scale = c(3.5, 0.8)
    )
  )
```

::: {.callout-note}
A word cloud is useful as a quick visual impression, but a bar chart is often easier to read and compare.
:::

# Part 7 — Data from multiple sources

## Lab 22 — Why joins?

Real research data often come from different tables:

- students and projects;
- patients and hospital records;
- firms and balance sheets;
- athletes and sensor measurements;
- documents and metadata.

A join is not just a technical operation: it can change the population being analysed.

## Lab 23 — Two related tables

```{r}
students <- tibble::tribble(
  ~student_id, ~name, ~programme,
  1, "Anna", "Engineering",
  2, "Marco", "Education/Sport",
  3, "Sara", "Economics/Management",
  4, "Luca", "Humanities"
)

projects <- tibble::tribble(
  ~student_id, ~topic,
  1, "Sensor data",
  2, "Training load",
  3, "Firm innovation",
  5, "Archival sources"
)

students
projects
```

## Lab 24 — Inner join

An inner join keeps only records that match in both tables.

```{r}
students |>
  inner_join(projects, by = "student_id")
```

::: {.callout-tip}
### Your turn

Which student is lost after the inner join?  
Why?
:::

Write your answer here:

```{markdown}
Answer:

```

## Lab 25 — Left join

A left join keeps all records from the first table and adds matching information from the second table.

```{r}
students |>
  left_join(projects, by = "student_id")
```

::: {.callout-tip}
### Your turn

Which student has missing project information?
:::

Write your answer here:

```{markdown}
Answer:

```

## Lab 26 — Anti joins

Anti joins find unmatched records.

```{r}
students |>
  anti_join(projects, by = "student_id")
```

```{r}
projects |>
  anti_join(students, by = "student_id")
```

::: {.callout-important}
Different joins answer different research questions. Always ask which observations are kept and which are lost.
:::

# Final reflection

Write a short paragraph answering the following questions:

1. What did you learn about visualising one variable?
2. What did you learn about visualising relationships?
3. What did you learn about transforming text into data?
4. What did you learn about joins?

```{markdown}
Final reflection:

```
