---
title: "Essential Statistical Data Analysis"
subtitle: "Class 2 Lab — Classification: from probabilities to decisions"
author: "YOUR NAME GOES HERE"
date: "May 18, 2026"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    code-tools: true
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false

library(patchwork)
library(tidyverse)
library(tidymodels)
library(kableExtra)
library(palmerpenguins)

# Some classifiers used later in the lab are available through additional packages.
# If they are not installed, the optional sections will be skipped.
has_discrim <- requireNamespace("discrim", quietly = TRUE)
has_MASS <- requireNamespace("MASS", quietly = TRUE)
has_naivebayes <- requireNamespace("naivebayes", quietly = TRUE)
has_nnet <- requireNamespace("nnet", quietly = TRUE)

if (has_discrim) {
  library(discrim)
}

theme_set(theme_minimal())
```

# Lab overview

In this lab, we practise the main steps of a classification analysis:

- identifying a categorical response variable;
- splitting data into training and test sets;
- fitting a classification model;
- obtaining predicted probabilities;
- converting probabilities into predicted classes;
- evaluating predictions with confusion matrices;
- changing the classification threshold;
- drawing ROC curves and computing AUC;
- comparing different classifiers.

You can edit the code chunks directly and re-run them.

# Part 1 — The classification problem

We use the `penguins` dataset from the `palmerpenguins` package.

Each row describes one penguin. The variables include species, island, bill measurements, flipper length, body mass, sex, and year.

To focus on binary classification, we create a new response variable:

$$
\texttt{adelie} =
\begin{cases}
\text{Yes}, & \text{if the penguin is Adelie} \\
\text{No}, & \text{otherwise}
\end{cases}
$$

The goal is to predict whether a penguin is an Adelie penguin from its measurements.

## Lab 1 — Load and prepare the data

```{r}
penguins_binary <- palmerpenguins::penguins |>
  drop_na(
    species,
    bill_length_mm,
    bill_depth_mm,
    flipper_length_mm,
    body_mass_g,
    island,
    sex
  ) |>
  mutate(
    adelie = factor(
      if_else(species == "Adelie", "Yes", "No"),
      levels = c("No", "Yes")
    ),
    species = as.factor(species),
    island = as.factor(island),
    sex = as.factor(sex)
  )

glimpse(penguins_binary)
```

::: {.callout-tip}
### Your turn

Identify:

1. the observational unit;
2. the response variable;
3. the predictor variables we could use.
:::

Write your answer here:

```{markdown}
Observational unit:

Response variable:

Possible predictor variables:
```

## Lab 2 — Class frequencies

Before fitting a classifier, we should inspect the distribution of the response variable.

```{r}
penguins_binary |>
  count(adelie)
```

```{r}
penguins_binary |>
  count(adelie) |>
  mutate(proportion = n / sum(n))
```

::: {.callout-tip}
### Your turn

Is this a balanced or imbalanced classification problem?  
Which class is more frequent?
:::

Write your answer here:

```{markdown}
Answer:
```

# Part 2 — Exploring predictors and classes

## Lab 3 — Bill length and Adelie status

```{r}
penguins_binary |>
  ggplot(aes(x = adelie, y = bill_length_mm, fill = adelie)) +
  geom_boxplot(alpha = 0.6, show.legend = FALSE) +
  labs(
    x = "Adelie",
    y = "Bill length (mm)"
  )
```

::: {.callout-tip}
### Your turn

Does `bill_length_mm` look useful for predicting whether a penguin is Adelie?  
Why?
:::

Write your interpretation here:

```{markdown}
Interpretation:
```

## Lab 4 — Bill depth and Adelie status

```{r}
penguins_binary |>
  ggplot(aes(x = adelie, y = bill_depth_mm, fill = adelie)) +
  geom_boxplot(alpha = 0.6, show.legend = FALSE) +
  labs(
    x = "Adelie",
    y = "Bill depth (mm)"
  )
```

::: {.callout-tip}
### Your turn

Does `bill_depth_mm` look useful for distinguishing Adelie penguins from other penguins?
:::

Write your interpretation here:

```{markdown}
Interpretation:
```

## Lab 5 — Two predictors at once

```{r}
penguins_binary |>
  ggplot(aes(x = bill_length_mm, y = bill_depth_mm, colour = adelie)) +
  geom_point(alpha = 0.7) +
  labs(
    x = "Bill length (mm)",
    y = "Bill depth (mm)",
    colour = "Adelie"
  )
```

::: {.callout-tip}
### Your turn

What do you notice about the position of Adelie penguins in the plot?
:::

Write your interpretation here:

```{markdown}
Interpretation:
```

# Part 3 — Training and test data

## Lab 6 — Split the data

We split the data into:

- a training set, used to fit the model;
- a test set, used to evaluate predictions on new observations.

```{r}
set.seed(1234)

peng_split <- initial_split(
  penguins_binary,
  prop = 3/4,
  strata = adelie
)

peng_train <- training(peng_split)
peng_test <- testing(peng_split)
```

```{r}
peng_train |>
  count(adelie) |>
  mutate(proportion = n / sum(n))
```

```{r}
peng_test |>
  count(adelie) |>
  mutate(proportion = n / sum(n))
```

::: {.callout-tip}
### Your turn

Why did we use `strata = adelie` when splitting the data?
:::

Write your answer here:

```{markdown}
Answer:
```

# Part 4 — Logistic regression

## Lab 7 — Recipe, model, workflow

We first fit a logistic regression with one predictor: `bill_length_mm`.

```{r}
peng_rec <- recipe(adelie ~ bill_length_mm, data = peng_train)

peng_log_spec <- logistic_reg(
  mode = "classification",
  engine = "glm"
)

peng_log_wflow <- workflow() |>
  add_recipe(peng_rec) |>
  add_model(peng_log_spec)

peng_log_fit <- peng_log_wflow |>
  fit(data = peng_train)
```

::: {.callout-note}
The workflow has three ingredients: recipe, model specification, and fitted model.
:::

## Lab 8 — Look at the fitted model

```{r}
peng_log_fit |>
  tidy()
```

::: {.callout-tip}
### Your turn

Look at the sign of the coefficient for `bill_length_mm`.  
Does the probability of being Adelie increase or decrease when bill length increases?
:::

Write your answer here:

```{markdown}
Answer:
```

## Lab 9 — Predicted classes and probabilities

A classifier can return:

- a predicted class;
- predicted probabilities for each class.

```{r}
log_pred <- peng_log_fit |>
  augment(new_data = peng_test)

log_pred |>
  select(adelie, bill_length_mm, .pred_class, .pred_No, .pred_Yes) |>
  slice_head(n = 10) |>
  kbl() |>
  kable_styling(full_width = FALSE)
```

::: {.callout-tip}
### Your turn

For the first 10 observations, find one observation with relatively high predicted probability of being Adelie.
:::

Write your answer here:

```{markdown}
Answer:
```

# Part 5 — From probabilities to decisions

## Lab 10 — Default threshold

By default, predicted probabilities are usually converted into classes using threshold 0.5:

$$
\widehat{P}(\text{Adelie} = \text{Yes} \mid X) > 0.5
$$

```{r}
log_pred <- log_pred |>
  mutate(
    pred_0_5 = factor(
      if_else(.pred_Yes > 0.5, "Yes", "No"),
      levels = levels(peng_test$adelie)
    )
  )

log_pred |>
  select(adelie, .pred_Yes, .pred_class, pred_0_5) |>
  slice_head(n = 10)
```

::: {.callout-tip}
### Your turn

Are `.pred_class` and `pred_0_5` the same for the first 10 observations?
:::

Write your answer here:

```{markdown}
Answer:
```

## Lab 11 — Confusion matrix

```{r}
log_pred |>
  conf_mat(truth = adelie, estimate = .pred_class)
```

```{r}
log_pred |>
  conf_mat(truth = adelie, estimate = .pred_class) |>
  autoplot(type = "heatmap")
```

::: {.callout-tip}
### Your turn

How many actual Adelie penguins were correctly predicted as Adelie?
:::

Write your answer here:

```{markdown}
Answer:
```

## Lab 12 — Accuracy, sensitivity and specificity

```{r}
log_pred |>
  metrics(truth = adelie, estimate = .pred_class)
```

```{r}
log_pred |>
  sens(truth = adelie, estimate = .pred_class, event_level = "second")
```

```{r}
log_pred |>
  spec(truth = adelie, estimate = .pred_class, event_level = "second")
```

::: {.callout-note}
Here the event of interest is `Yes`, so we use `event_level = "second"` because the factor levels are `No`, `Yes`.
:::

::: {.callout-tip}
### Your turn

Is the model better at identifying Adelie penguins or non-Adelie penguins?
:::

Write your answer here:

```{markdown}
Answer:
```

# Part 6 — Changing the threshold

## Lab 13 — Several thresholds

In some applications, the threshold 0.5 is not the best choice.

For example, if failing to identify an Adelie penguin is costly, we may want to use a lower threshold.

```{r}
log_pred_thresholds <- log_pred |>
  mutate(
    pred_0_20 = factor(if_else(.pred_Yes > 0.20, "Yes", "No"), levels = levels(penguins_binary$adelie)),
    pred_0_30 = factor(if_else(.pred_Yes > 0.30, "Yes", "No"), levels = levels(penguins_binary$adelie)),
    pred_0_40 = factor(if_else(.pred_Yes > 0.40, "Yes", "No"), levels = levels(penguins_binary$adelie)),
    pred_0_50 = factor(if_else(.pred_Yes > 0.50, "Yes", "No"), levels = levels(penguins_binary$adelie)),
    pred_0_60 = factor(if_else(.pred_Yes > 0.60, "Yes", "No"), levels = levels(penguins_binary$adelie))
  )
```

```{r}
threshold_summary <- log_pred_thresholds |>
  pivot_longer(
    cols = starts_with("pred_0_"),
    names_to = "threshold",
    values_to = "prediction"
  ) |>
  group_by(threshold) |>
  summarise(
    accuracy = mean(adelie == prediction),
    sensitivity = sens_vec(
      truth = adelie,
      estimate = prediction,
      event_level = "second"
    ),
    specificity = spec_vec(
      truth = adelie,
      estimate = prediction,
      event_level = "second"
    ),
    .groups = "drop"
  )

threshold_summary |>
  kbl(digits = 3) |>
  kable_styling(full_width = FALSE)
```

::: {.callout-tip}
### Your turn

What happens to sensitivity when the threshold decreases?  
What happens to specificity?
:::

Write your interpretation here:

```{markdown}
Interpretation:
```

## Lab 14 — Confusion matrices for different thresholds

```{r, fig.align='center', out.width="70%"}
p_20 <- log_pred_thresholds |>
  conf_mat(truth = adelie, estimate = pred_0_20) |>
  autoplot(type = "heatmap") +
  ggtitle("threshold = 0.20")

p_40 <- log_pred_thresholds |>
  conf_mat(truth = adelie, estimate = pred_0_40) |>
  autoplot(type = "heatmap") +
  ggtitle("threshold = 0.40")

p_60 <- log_pred_thresholds |>
  conf_mat(truth = adelie, estimate = pred_0_60) |>
  autoplot(type = "heatmap") +
  ggtitle("threshold = 0.60")

p_20 | p_40 | p_60
```

::: {.callout-tip}
### Your turn

Which threshold would you choose if missing an Adelie penguin is very costly?  
Why?
:::

Write your answer here:

```{markdown}
Answer:
```

# Part 7 — ROC curve and AUC

## Lab 15 — ROC curve

```{r}
log_pred |>
  roc_curve(truth = adelie, .pred_Yes, event_level = "second") |>
  ggplot(aes(x = 1 - specificity, y = sensitivity)) +
  geom_path() +
  geom_abline(lty = 3) +
  coord_equal() +
  labs(
    title = "Logistic regression ROC curve",
    x = "1 - specificity",
    y = "Sensitivity"
  )
```

## Lab 16 — AUC

```{r}
log_pred |>
  roc_auc(truth = adelie, .pred_Yes, event_level = "second")
```

::: {.callout-tip}
### Your turn

What does a larger AUC mean?
:::

Write your answer here:

```{markdown}
Answer:
```

# Part 8 — Adding more predictors

## Lab 17 — Multiple logistic regression

We now use several numerical predictors: bill measurements, flipper length, and body mass.

```{r}
peng_rec_multi <- recipe(
  adelie ~ bill_length_mm + bill_depth_mm + flipper_length_mm + body_mass_g,
  data = peng_train
)

peng_log_wflow_multi <- workflow() |>
  add_recipe(peng_rec_multi) |>
  add_model(peng_log_spec)

peng_log_fit_multi <- peng_log_wflow_multi |>
  fit(data = peng_train)

log_pred_multi <- peng_log_fit_multi |>
  augment(new_data = peng_test)
```

```{r}
log_pred_multi |>
  roc_auc(truth = adelie, .pred_Yes, event_level = "second")
```

::: {.callout-tip}
### Your turn

Compare the AUC of the simple model with the AUC of the multiple-predictor model.  
Did the additional predictors help?
:::

Write your answer here:

```{markdown}
Answer:
```

## Lab 18 — Probability curve

The simple model uses only `bill_length_mm`, so we can easily visualise how the predicted probability changes with bill length.

```{r}
new_penguins <- tibble(
  bill_length_mm = seq(
    from = min(peng_train$bill_length_mm),
    to = max(peng_train$bill_length_mm),
    length.out = 200
  )
)

pred_grid <- peng_log_fit |>
  augment(new_data = new_penguins)

pred_grid |>
  ggplot(aes(x = bill_length_mm, y = .pred_Yes)) +
  geom_line(linewidth = 1) +
  labs(
    x = "Bill length (mm)",
    y = "Predicted probability of being Adelie"
  )
```

::: {.callout-tip}
### Your turn

How does the estimated probability of being Adelie change as bill length increases?
:::

Write your interpretation here:

```{markdown}
Interpretation:
```

# Part 9 — Other classifiers

## Lab 19 — Linear discriminant analysis

This section requires the `discrim` and `MASS` packages.

```{r}
#| eval: !expr has_discrim && has_MASS

peng_rec_discrim <- recipe(
  adelie ~ bill_length_mm + bill_depth_mm + flipper_length_mm + body_mass_g,
  data = peng_train
)

peng_lda_spec <- discrim_linear(
  mode = "classification",
  engine = "MASS"
)

peng_lda_wflow <- workflow() |>
  add_recipe(peng_rec_discrim) |>
  add_model(peng_lda_spec)

peng_lda_fit <- peng_lda_wflow |>
  fit(data = peng_train)

lda_pred <- peng_lda_fit |>
  augment(new_data = peng_test)

lda_pred |>
  roc_auc(truth = adelie, .pred_Yes, event_level = "second")
```

::: {.callout-note}
LDA is a generative classifier. It assumes that predictors are approximately normally distributed within each class and that classes share the same covariance structure.
:::

## Lab 20 — Quadratic discriminant analysis

This section requires the `discrim` and `MASS` packages.

```{r}
#| eval: !expr has_discrim && has_MASS

peng_qda_spec <- discrim_quad(
  mode = "classification",
  engine = "MASS"
)

peng_qda_wflow <- workflow() |>
  add_recipe(peng_rec_discrim) |>
  add_model(peng_qda_spec)

peng_qda_fit <- peng_qda_wflow |>
  fit(data = peng_train)

qda_pred <- peng_qda_fit |>
  augment(new_data = peng_test)

qda_pred |>
  roc_auc(truth = adelie, .pred_Yes, event_level = "second")
```

::: {.callout-note}
QDA relaxes the common covariance assumption. This allows more flexible, curved decision boundaries.

QDA can fail when predictors are highly redundant or when the class-specific covariance matrix cannot be estimated reliably.
:::

## Lab 21 — Naive Bayes

This section requires the `discrim` and `naivebayes` packages.

```{r}
#| eval: !expr has_discrim && has_naivebayes

peng_nb_spec <- naive_Bayes(
  mode = "classification",
  engine = "naivebayes"
)

peng_nb_wflow <- workflow() |>
  add_recipe(peng_rec_discrim) |>
  add_model(peng_nb_spec)

peng_nb_fit <- peng_nb_wflow |>
  fit(data = peng_train)

nb_pred <- peng_nb_fit |>
  augment(new_data = peng_test)

nb_pred |>
  roc_auc(truth = adelie, .pred_Yes, event_level = "second")
```

::: {.callout-note}
Naive Bayes assumes that predictors are independent within each class. The assumption is strong, but the method can work well in practice.
:::

## Lab 22 — Compare classifiers

```{r}
#| eval: !expr has_discrim && has_MASS && has_naivebayes

roc_all <- bind_rows(
  log_pred_multi |>
    roc_curve(truth = adelie, .pred_Yes, event_level = "second") |>
    mutate(model = "Logistic regression"),

  lda_pred |>
    roc_curve(truth = adelie, .pred_Yes, event_level = "second") |>
    mutate(model = "LDA"),

  qda_pred |>
    roc_curve(truth = adelie, .pred_Yes, event_level = "second") |>
    mutate(model = "QDA"),

  nb_pred |>
    roc_curve(truth = adelie, .pred_Yes, event_level = "second") |>
    mutate(model = "Naive Bayes")
)

roc_all |>
  ggplot(aes(x = 1 - specificity, y = sensitivity, colour = model)) +
  geom_path(linewidth = 1, alpha = 0.8) +
  geom_abline(lty = 3) +
  coord_equal() +
  labs(
    title = "ROC curves for different classifiers",
    x = "1 - specificity",
    y = "Sensitivity",
    colour = "Model"
  )
```

```{r}
#| eval: !expr has_discrim && has_MASS && has_naivebayes

auc_all <- bind_rows(
  log_pred_multi |>
    roc_auc(truth = adelie, .pred_Yes, event_level = "second") |>
    mutate(model = "Logistic regression"),

  lda_pred |>
    roc_auc(truth = adelie, .pred_Yes, event_level = "second") |>
    mutate(model = "LDA"),

  qda_pred |>
    roc_auc(truth = adelie, .pred_Yes, event_level = "second") |>
    mutate(model = "QDA"),

  nb_pred |>
    roc_auc(truth = adelie, .pred_Yes, event_level = "second") |>
    mutate(model = "Naive Bayes")
)

auc_all |>
  select(model, .estimate) |>
  arrange(desc(.estimate)) |>
  kbl(digits = 3) |>
  kable_styling(full_width = FALSE)
```

::: {.callout-tip}
### Your turn

Which classifier has the largest AUC?  
Is the difference large enough to matter in practice?
:::

Write your answer here:

```{markdown}
Answer:
```

# Optional — Multi-class classification

So far, we transformed the problem into a binary classification task.

We can also predict the original `species` variable.  
This is a multi-class classification problem.

## Lab 23 — Prepare the multi-class data

```{r}
penguins_multi <- palmerpenguins::penguins |>
  drop_na(
    species,
    bill_length_mm,
    bill_depth_mm,
    flipper_length_mm,
    body_mass_g
  ) |>
  mutate(species = as.factor(species))

set.seed(1234)

peng_multi_split <- initial_split(
  penguins_multi,
  prop = 3/4,
  strata = species
)

peng_multi_train <- training(peng_multi_split)
peng_multi_test <- testing(peng_multi_split)
```

## Lab 24 — Fit a multinomial logistic regression

```{r}
#| eval: !expr has_nnet

peng_multi_rec <- recipe(
  species ~ bill_length_mm + bill_depth_mm + flipper_length_mm + body_mass_g,
  data = peng_multi_train
)

peng_multi_spec <- multinom_reg(
  mode = "classification",
  engine = "nnet"
)

peng_multi_wflow <- workflow() |>
  add_recipe(peng_multi_rec) |>
  add_model(peng_multi_spec)

peng_multi_fit <- peng_multi_wflow |>
  fit(data = peng_multi_train)

peng_multi_pred <- peng_multi_fit |>
  augment(new_data = peng_multi_test)

peng_multi_pred |>
  conf_mat(truth = species, estimate = .pred_class)
```

::: {.callout-tip}
### Your turn

Which species are easiest to classify?  
Which species are most often confused?
:::

Write your answer here:

```{markdown}
Answer:
```

# Final reflection

Write a short paragraph answering the following questions:

1. What is the difference between a predicted probability and a predicted class?
2. Why can changing the threshold change the type of errors we make?
3. Why is accuracy sometimes misleading in imbalanced classification problems?
4. Which part of the classification workflow would you use in your own research?

```{markdown}
Final reflection:
```
